Red Wine Quality by Neal Kfoury
========================================================
```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.
library(ggplot2)
library(GGally)
library(dplyr)
library(ggcorrplot)
```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
getwd()
setwd('~/Desktop/more\ of\ zinger\'s\ things/DAND/Unit\ 5')

wineInfo <- read.csv('wineQualityReds.csv')
```

In this project, I will explore a dataset of red wines containing information
about several chemical features of wines. Each entry in the dataset also
includes a quality rating as assessed by wine experts. Hopefully I will be able
to uncover some promising correlations between certain chemical features of red
wines and the quality ratings of those wines.

First, let's take a look at the structure of the dataset:

```{r echo=FALSE, message=FALSE, warning=FALSE, introduction_structure}
str(wineInfo)
```

We can see that our dataset contains 1599 observations (rows) of 13 variables
(columns), and that all our data is numerical, and looks quite tidy. X
represents the ID of each wine as a nominal int, and quality represents the
interval rating of a given wine as an int (according to wine experts).

Since our data is numerical, we should be able to easily analyze it with common
data analysis techniques. Let's take a closer look at each of the columns:

```{r echo=FALSE, message=FALSE, warning=FALSE, introduction_summary}
summary(wineInfo)
```

Here we can see some descriptive statistics for each of our 19 columns in the
dataset. 

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(aes(x = quality), data = wineInfo) +
  geom_bar(width = 0.9) +
  scale_x_continuous(breaks = 3:8)
```

Since I'm most interested in comparing my variables against wine quality, I'm
interested in seeing the distribution of wine quality ratings in the dataset.
Although the dataset documentation claims the wines are rated from 0-10, our
dataframe summary revealed the range to fall between 3-8 in our sample. Since
the quality ratings are categorical, I used a bar chart. It looks like most
wines weigh in at 5-6, according to the experts.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wineInfo, aes(x = chlorides)) +
  geom_histogram(binwidth = 0.01)

ggplot(data = wineInfo, aes(x = chlorides)) +
  geom_histogram(binwidth = 0.01) +
  scale_x_log10()

summary(wineInfo$chlorides)
```

Looking at the statistical summary of my columns, chlorides was the first column
I noticed with a suspiscious-looking distribution: a mean much higher than the 
median, and a maximum much higher than the 3rd quartile. Plotting chlorides
resulted in a very skewed chart. I transformed it on a log10 scale to find a
normal-ish distribution.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wineInfo, aes(x = free.sulfur.dioxide)) +
  geom_histogram(binwidth = 1)

ggplot(data = wineInfo, aes(x = total.sulfur.dioxide)) +
  geom_histogram(binwidth = 3)

ggplot(data = wineInfo, aes(x = sulphates)) +
  geom_histogram(binwidth = 0.05)
```

I suspect that total sulfur dioxide and free sulfur dioxide may be correlated,
and they do seem to have similar looking distributions. They also seem to share
a shape with sulphates. I'll keep this in mind.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wineInfo, aes(x = residual.sugar)) +
  geom_histogram(bindwidth = 0.3)

ggplot(data = wineInfo, aes(x = residual.sugar)) +
  geom_histogram(binwidth = 0.03) + 
  scale_x_log10()

summary(wineInfo$residual.sugar)
```

I applied a log10 transform for residual sugar, but I am still seeing a long 
tail. There seems to be quite a lot of variance in this feature.

```{r echo=FALSE, message=FALSE, warning=FALSE}
qplot(x =fixed.acidity, data = wineInfo)
qplot(x = volatile.acidity, data = wineInfo)
qplot(x = citric.acid, data = wineInfo)
qplot(x = pH, data = wineInfo)
```

Some quick plots of the acidity-related features of the dataset (fixed acidity,
volatile acidity, citric acid, pH). As with the sulfur dioxides, it seems like
fixed acidity and volatile acidity may be correlated. pH looks like it has a
beautiful normal distribution. I'll keep in mind that pH is itself a log scale.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wineInfo, aes(x = density)) +
  geom_histogram(binwidth = 0.0005)
```

Finally, a nice, normal distribution. The dataset notes say density is
represented as g/mL. Thinking back to high school chemistry class, I remember 
that density is a physical property, rather than a chemical property. It might
not be strongly correlated with any other individual feature.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wineInfo, aes(x = alcohol)) +
  geom_histogram(bindwidth = 0.1)

ggplot(data = wineInfo, aes(x = alcohol)) +
  geom_histogram(binwidth = 0.01) + 
  scale_x_log10()

summary(wineInfo$alcohol)
```

I am again surprised that a log10 transform did not seem to change the general
shape of the tail. I recognize this shape from some previous plots (residual 
sugar, e.g.). Maybe I will see some important correlations with alcohol. 

# Univariate Analysis

### What is the structure of your dataset?
My dataset includes 11 chemical features of 1599 red wines, plus an ID and
quality classification for each.

### What is/are the main feature(s) of interest in your dataset?
My main insterest is the "output" feature of the dataset--quality. This feature 
is a subjective rating, assigned by wine experts, and ranges between 0-10;
however, in my dataset, I only have instances ranging from 3-8, with most 
falling from 5-6.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?
I am interested in investigating the relationship between sugar and alcohol, 
since their plots have similar shapes. It will also be important to confirm my
suspiscions about the relationships between chemically-similar features (such as
those all pertaining to sulfur, or acidity).

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?
I applied log transforms to some plots in order to better understand the form
of the long right-skewed tails that appeared in many of the linear plots.

# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots}
# Calculate correlation coefficients for all pairs of features
corr <- round(cor(wineInfo[2:13]),3)
ggcorrplot(corr, method = 'circle')
```

Above I have plotted a matrix of each possible bivariate scatter amongst the
non-X columns in the dataset. I can get a general sesnse for interesting
correlations between variables in the data.

```{r echo=FALSE, message=FALSE, warning=FALSE}
corr
```

Here I can see the numerical representation of the correlations. At a glance,
quality is most strongly correlated with volatile acidity (negative
correlation), citric acid, sulphates, and alcohol (positive correlations). I'll
want to be sure to explore those in more detail.

I'm also noticing that I spoke too soon about density--it's showing off
one of the strongest correlations in the matrix (0.668 with fixed.acidity). It
is also residual sugar's strongest correlative feature (0.355).

Free sulfur dioxide and total sulfur dioxide appear to be strongly correlated.
This doesn't come as much of a surprise.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(wineInfo, aes(x = quality, y = volatile.acidity)) +
  geom_point()

ggplot(wineInfo, aes(x = quality, y = volatile.acidity)) +
  geom_jitter(alpha = 0.3) +
  geom_smooth(method = "lm")

# Group the data by quality rating to summarize each group by feature.
quality_groups <- group_by(wineInfo, quality)

summarise(quality_groups,
          volatile_acidity_mean = mean(volatile.acidity),
          volatile_acidity_median = median(volatile.acidity),
          n = n())
```

Adding some jitter shows us a slightly clearer picture of the middle-grade
quality wines. Looking at the jitter, I can see a visual trend in the points
that matches the line of best fit.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(wineInfo, aes(x = quality, y = citric.acid)) +
  geom_jitter(alpha = 0.3)

summarise(quality_groups,
          citric_acid_mean = mean(citric.acid),
          citric_acid_median = median(citric.acid),
          n = n())
```

The citric acid feature includes a lot of 0 values, and I can easily see them
grouped at the bottom of the chart. This explains why the median citric acidity
of each quality group is more stable than the median. I am a little dubious of
the usefulness of this correlation for predicting wine quality.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(wineInfo, aes(x = quality, y = sulphates)) +
  geom_jitter(alpha = 0.3) +
  geom_smooth(method = "lm")

ggplot(wineInfo, aes(x = quality, y = sulphates, group = quality)) +
  geom_boxplot()
```

The dense bands in the jitter plot seemed to be dominated by the high-outlying
points, so I plotted them as a box and whisker to get a clearer picture of the
trend. Recall that this was the second strongest positive quality correlation in
the original matrix.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(wineInfo, aes(x = quality, y = alcohol)) +
  geom_jitter(alpha = 0.3) +
  geom_smooth(method = "lm")

summarise(quality_groups,
          citric_acid_mean = mean(alcohol),
          citric_acid_median = median(alcohol),
          n = n())
```

Although the higher quality wines (7, 8) do not contain many datapoints, it is
clear that the heavy bands of sub-10% alcohol content are concentrated in the
mid-grade wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(wineInfo, aes(x = density, y = fixed.acidity)) +
  geom_point() +
  geom_smooth(method = "lm")

# Subset the data according to residual sugar outliers (>=10 omitted).
sugar <- subset(wineInfo, residual.sugar < 10,
                # The features I will compare using this subset.
                select = c(residual.sugar, fixed.acidity, density, alcohol,
                           quality))

ggplot(sugar, aes(x = density, y = residual.sugar)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Here I'm taking a look at two of the promising correlations with density: fixed
acidity and residual sugar. Fixed acidity shows a pretty clear positive
correlation. The residual sugar plot contained a handful of outliers--I removed
those wines with a residual sugar of 10+ (11 rows). Even without those outliers
gone, the sugar correlation with density seems shallow.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?
The main feature of interest (quality) showed promising correlations with
alcohol, volatile acidity, and sulphates.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?
According to the dataset documentation, the density property depends on alcohol
and residual sugar content. Interestingly, density was fairly strongly 
correlated with quite a few features of the data--fixed acidity and citric acid,
in addition to residual sugar and alcohol. Since density seems to correlate 
with so many features, I was surprised to find that it did not have a strong
relationship with quality.

### What was the strongest relationship you found?
The strongest relationships in the data were among the more obviously-related
features: free sulphates with total sulphates, and citric acid with fixed
acidity, for example.

In my analysis I am most interested in quality. Quality was most strongly
related with alcohol content (positively correlated) and volatile acidity
(negatively correlated).

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
ggplot(wineInfo, aes(x = alcohol, y = density, color = quality)) +
  geom_jitter() +
  scale_color_gradient(high = 'maroon', low = 'white')
```

This plot appears to show higher quality wines trending towards the bottom-left
corner (higher alcohol, lower density). The correlation seems very slight.

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(sugar, aes(y = alcohol, x = log2(residual.sugar), color = quality)) +
  geom_jitter() +
  scale_color_gradient(high = 'yellow', low = 'blue')
```

This plot compares quality to alcohol and residual sugar. I applied a log2
transform to get a better view, since the sugar distribution from the univariate
analysis had such a long tail.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plot on exp scale to view pH linearly.
ggplot(wineInfo, aes(x = fixed.acidity, y = citric.acid,
                     color = exp(pH))) +
  geom_point() +
  scale_color_gradient(high = 'blue', low = 'red')
```

Although seems to drive fixed acidity towards an asymptote, pH is a good 
indicator of both citric acid and fixed acidity. Since pH is itself a log scale,
I represented it here as an exponent function.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Subset the data by removing the top 10% of total.sulfur.dioxide.
sulphates_trimmed <- subset(wineInfo,
                    total.sulfur.dioxide < quantile(total.sulfur.dioxide, 0.9),
                    # The features I will compare with this subset.
                    select = c(sulphates, total.sulfur.dioxide,
                               free.sulfur.dioxide, volatile.acidity,
                               quality))

ggplot(sulphates_trimmed, aes(x = total.sulfur.dioxide, y = free.sulfur.dioxide,
                     color = sulphates)) +
  geom_point() +
  scale_color_gradient(high = 'red', low = 'blue')
```

During the bivariate exploration phase, I noticed that sulphates did not show
a strong correlation with free SO2 or total SO2 (although the latter two were
correlated with one another). I plotted the three together just to see what was
going on. To get a better picture (without outliers), I scaled out the top 10%
of total sulfur dioxide. The result seems to confirm that the high sulphate
wines are scattered amongst the distribution of total/free sulfur dioxide wines.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Subset the data by removing the top 10% of sulphates.
sulphates_trimmed2 <- subset(wineInfo,
                            sulphates < quantile(sulphates, 0.9),
                            select = c(sulphates, alcohol, quality))

ggplot(sulphates_trimmed2, aes(x = alcohol, y = sulphates, color = quality)) +
  geom_jitter(alpha = 0.8) +
  scale_color_gradient(high = 'yellow', low = 'blue')

ggplot(wineInfo, aes(x = alcohol, y = volatile.acidity, color = quality)) +
  geom_jitter(alpha = 0.8) +
  scale_color_gradient(high = 'yellow', low = 'blue')

ggplot(sulphates_trimmed, aes(x = sulphates, y = volatile.acidity,
                              color = quality)) +
  geom_jitter(alpha = 0.8) +
  scale_color_gradient(high = 'yellow', low = 'blue')
```

I know that alcohol is somewhat correlated with quality. I chose to plot and
compare that relationship against two other strong quality correlates: sulphates
(positive quality correlation) and volatile acidity (negative quality
correlation). As expected, high quality wines of all alcohol contents trend
towards higher sulfates, and lower volatile acidity.

Finally, I took a look at sulphates compared to volatile acidity. It looks like
wines generally have high volatile acidities and low qualities at low sulphate
levels. I cut some outliers out of this last plot to see the large cluster more
clearly.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?
Sulphates and volatile acidity each independently strengthened the predictive
power of alcohol content on quality. For example, for low-alcohol wines 
(typically given lower quality ratings), it looks more likely to find a high
quality wine with a lower volatile acidity.

### Were there any interesting or surprising interactions between features?
Since density and alcohol were strong correlates of quality, I expected to see
them strengthen one another. However, plotting them together revealed caused the
relationship to become a little muddier. 

I was surprised to find little to no correlation between sulphates, total sulfur
dioxide, and free sulfur dioxide. I guess I should brush up on my chemistry.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
ggplot(aes(x = quality, alpha = 0.75), data = wineInfo) +
  geom_bar(width = 0.9) +
  scale_x_continuous(breaks = 3:8) +
  geom_line(y = 100, color = "blue") +
  annotate('text',
           x = 3.2,
           y = 130,
           label = paste('n=100',
                         sep = ''),
           color = 'blue') +
  geom_line(y = 30, color = "red") +
  annotate('text',
           x = 3.2,
           y = 60,
           label = paste('n=30',
                         sep = ''),
           color = 'red') +
  ggtitle("Number of Wines by Quality Rating") +
  xlab("Quality Rating") +
  ylab("Number of Wines")
```

### Description One
A simple bar chart representing the number of wine instances for each quality
rating in the dataset. I have included it to emphasize one of the main
limitations of the dataset: the number of samples. Although the dataset is
almost 1600 samples wide, most of those samples fall in the middle 2 categories.
There are no examples at all for quality groups 0, 1, 2, 9, and 10; and, the
relative dearth of examples in categories 3, 4, 7, and 8 seriously limits my
ability to draw meaningful conclusions when I compare quality to the other
features of the datset. 

I would feel more confident identifying trends in wine quality if the number of
samples for each quality rating was at least 30, and preferably 100 or more.
In the dataset I used in this project, fewer than half of the quality bins
meet this criterion.

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
ggplot(wineInfo, aes(x = quality, y = volatile.acidity, group = quality,)) +
  geom_boxplot() +
  geom_smooth(method = "lm") +
  ggtitle("Volatile Acidity by Wine Quality") +
  xlab("Quality Rating") +
  ylab("Volatile Aciditiy (g/L)")
```

### Description Two
I included this boxplot version of my jitter plot from the univariate analysis
phase, comparing quality and volatile acidity. The box plot makes it easier to
see the relationship: as quality increases, volatile acidity tends to decrease,
indicating that volatile acidity is an undesirable quality of red wines.

This plot would have been much more interesting if the dataset had included
more samples in the extreme quality ranges. With only a handful of samples at
either end, the original jitter plot was fairly hard to read, even with a line
of best fit overlaid. Additional datapoints would have allowed me to determine
whether the correlation is really as strong as it looks here.

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Three}
ggplot(sulphates_trimmed2, aes(y = alcohol, x = sulphates, color = factor(quality))) +
  geom_smooth(method = 'lm') +
  scale_color_brewer(type = "seq", palette = 3) +
  ggtitle("Quality by Alcohol and Sulphate Content") +
  ylab("Alcohol (% by volume)") +
  xlab("Sulphates (g/L)")
```

### Description Three
This plot compares the distribution of wine qualities across sulphates and 
alcohol content. I have re-plotted the scatter plot from the multivariate 
analysis phase as a line plot to better highlight the ranges for each quality
category across sulphates and alcohol content. I selected this plot in part
because of the clear relationship it shows between high quality and high alcohol
and sulphates content. My other reason is morepractical: I have heard that some
folks' bodies have a hard time processing sulphates. For a wine shopper with
such a condition, this plot would highlight the importance of selecting a high
(~12-13%) alcohol wine.

Additionally, this plot hightlights the relationship between sulphates and 
quality, especially for high-alcohol wines. For a wine with high alcohol 
content, high wine qualities trend somewhat toward higher sulphates content.
------

# Reflection
In the course of the analysis, I ran into difficulty trying to interpret plots
in which I compared 1 or more features against each of the quality ratings. Due
to a lack of data, most trends I identified in my analysis need to be heavily
extrapolated to wines in the more extreme quality categories. The 5 and 6 
quality ratings were so densely populated with samples (compared to the other
categories) that it was often quite difficult to visualize the trends between
them, much less to extend those trends to other quality categories. I was able 
to reduce the impact of this problem by making use of jitter and linear
regressions, which added some visual hints to my plots. Even so, some plots
(citric acid vs quality, e.g.) were simply impossible to read clearly.

That said, I did meet with some success when I was able to clearly isolate my
dataset's handful of 7 and 8 quality wines on one end of a multivariate
comparison, such as quality vs volatile acidity and alcohol content. This was
encouraging, as it is possible that new data might continue a clear-looking
trend. In particular, comparing alcohol content alongside key features like
sulphates and volatile acidity resulted in pretty decent quality trends, 
considering the limitations of the dataset.

If I had more data, I could easily have applied some of the same analyses, and
possibly used those findings to construct a good predictor of wine quality on
the basis of the chemical features included in this dataset.
